Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-06-08 talk/icooolps2011/paper.tex
fix the typos the reviewers pointed out

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have a rather complex core object
semantics. For them, even implementing just an interpreter is already a complex
task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project, which used the technique for dynamic
machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

@@

Since PyPy wants to be a general framework, we want to reuse our tracer for
different languages, which makes classical tracers inapplicable, because they
encode language semantics.
Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. In the model classes contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute on an instance, the instance's attributes are searched. If the
attribute is not found there, the class' methods are searched.

@@

When using this object model in
an interpreter, a large amount of time will be spent doing lookups in these
dictionaries.
Let's assume we trace through code that sums three attributes, such as:
\anto{I still think it's a bit weird to call them ``methods'' and then use
  them as attributes in the example}

@@

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace is in SSA form. Note how all the guards in trace correspond to a
condition in the original code. The trace contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups somehow. How to achieve this will be topic of
Section~\ref{sec:fastobjmodel}.

@@

Let's make this more concrete. If we trace a call to the function (written in
RPython) on the left, we get the trace on the right:

@@

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the RPython code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:

@@

The hint indicates that \texttt{x} is likely a runtime constant and the JIT
should try to perform runtime specialization on it
in the code that follows.\footnote{For technical reasons the promote hint needs
to be written down slightly differently in the actual code.}  When just running
the code, the \texttt{promote} function has no
effect. When tracing, some extra work
is done. Let's assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning.

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the runtime value of \texttt{x} as it was during tracing, which can
then be exploited by the compiler. The introduced
guard specializes the trace, because it only works if the value of $x_1$ is
\texttt{4}. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the first example. After the guard, the rest of the trace can
assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

@@

This new trace will be attached to the guard instruction of the first trace. If
$x_1$ takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that don't vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much assembler code being generated.

@@

This hint can be used to mark functions as \emph{trace-elidable}. A function is
termed trace-elidable if, during the execution of the program,
successive calls to the function with identical arguments always return the
same result. In addition the function needs to have no side effects or
idempotent side effects\footnote{This property
is less strict than that of a pure function, because it is only about actual
calls during execution. All pure functions are trace-elidable though.}.
From this definition follows that a call to an trace-elidable function with
constant arguments in a trace can be replaced with the result of the call seen during tracing.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{c}
is an trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
If the code in \texttt{c} should be constant-folded away, we would change the
class as follows:

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints of the previous section
don't seem to help with the current object model. There is
no trace-elidable function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Maps are shared between
different instances, therefore they have to be immutable, which means
that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map, which makes its side
effects idempotent. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to the
that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
memory address of the \texttt{Map} instance that has been promoted. Operations
that can be optimized away are grayed out, their results will be replaced with
fixed values by the constant folding.

@@

What we would really like is if the \texttt{Class.find\_method} method were trace-elidable.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
trace-elidable, because when the version object changes, the result of the call might be
different than the previous one.

@@

SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
used to trace through an JavaScript implementation written in C\#. The
JavaScript implementation compiles JavaScript to CIL bytecode together with an
implementation of the JavaScript object model. The object model uses maps
and inline caches to speed up operations on objects. The tracer traces through
the compiled JavaScript functions and the object model. SPUR contains two hints
that can be used to influence the tracer: one to prevent tracing of a C\#
function and one to force unrolling of a loop (PyPy has equivalent hints, but
they were not described in this paper).

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT.
%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

>>>>

One of the hardest parts of implementing an object-oriented dynamic language well is to
optimize its object model. This is made harder by the complexity of the core
object semantics of many recent languages such as Python, JavaScript or Ruby.
For them, even implementing just an interpreter is already a difficult
task. Implementing these languages efficiently with a just-in-time compiler (JIT) is
extremely challenging, because of their many corner-cases.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs were popularized by the Dynamo project, which used the technique for dynamic
machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

@@

Since PyPy wants to be a general framework, we want to reuse our tracer for
different languages, which makes classical tracers inapplicable, because they
encode language semantics.
Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces do not contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. In the model classes contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute of an instance, the instance's attributes are searched. If the
attribute is not found there, the class' methods are searched.

@@

When using this object model in
an interpreter, a large amount of time will be spent doing lookups in these
dictionaries.
Let us assume we trace through code that sums three attributes, such as:
\anto{I still think it's a bit weird to call them ``methods'' and then use
  them as attributes in the example}

@@

The trace would look like in Figure~\ref{fig:trace1}. In this example, the
attribute \texttt{a} is found on the instance, but the
attributes \texttt{b} and \texttt{c} are found on the class. The line
numbers in the trace correspond to the line numbers in
Figure~\ref{fig:interpreter-slow} where the traced operations come from. The
trace is in SSA form. Note how all the guards in trace correspond to a
condition in the original code. The trace contains
five calls to \texttt{dict.get}, which is slow. To make the language efficient
using a tracing JIT, we need to find a way to get rid of these dictionary
lookups. How to achieve this will be the topic of
Section~\ref{sec:fastobjmodel}.

@@

Let us make this more concrete. If we trace a call to the function (written in
RPython) on the left, we get the trace on the right:

@@

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let us assume that the value of \texttt{x} in the RPython code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:

@@

The hint indicates that \texttt{x} is likely a runtime constant and the JIT
should try to perform runtime specialization on it
in the code that follows.\footnote{For technical reasons the promote hint needs
to be written down slightly differently in the actual code.}  When just running
the code, the \texttt{promote} function has no
effect. When tracing, some extra work
is done. Let us assume that this changed function is traced with
the arguments \texttt{4} and \texttt{8}. The trace will be the same, except for one
operation at the beginning.

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the runtime value of \texttt{x} as it was during tracing, which can
then be exploited by the compiler. The introduced
guard specializes the trace, because it only works if the value of $x_1$ is
\texttt{4}. From the point of view of the
optimizer, this guard is not different frome the one produced by the \texttt{if}
statement in the first example. After the guard, it can be assumed that $x_1$
is equal to \texttt{4}, meaning that the optimizer will turn this trace into:

@@

This new trace will be attached to the guard instruction of the first trace. If
$x_1$ takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that do not vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much machine code being generated.

@@

This hint can be used to mark functions as \emph{trace-elidable}. A function is
termed trace-elidable if, during the execution of the program,
successive calls to the function with identical arguments always return the
same result. In addition the function needs to have no side effects or
idempotent side effects\footnote{This property
is less strict than that of a pure function, because it is only about actual
calls during execution. All pure functions are trace-elidable though.}.
From this definition follows that a call to a trace-elidable function with
constant arguments in a trace can be replaced with the result of the call seen during tracing.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{c}
is a trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
If the code in \texttt{c} should be constant-folded away, we would change the
class as follows:

@@

The first step in making \texttt{getattr} faster in our object model is to optimize
away the dictionary lookups on the instances. The hints of the previous section
do not seem to help with the current object model. There is
no trace-elidable function to be seen, and the instance is not a candidate for promotion,
because there tend to be many instances.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Maps are shared between
different instances, therefore they have to be immutable, which means
that their \texttt{getindex} method is a trace-elidable function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map, which makes its side
effects idempotent. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this adapted instance implementation, the trace we saw in Section~\ref{sub:running} changes to
that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
memory address of the \texttt{Map} instance that has been promoted. Operations
that can be optimized away are grayed out, their results will be replaced with
fixed values by the constant folding.

@@

What we would really like that the \texttt{Class.find\_method} method is trace-elidable.
But it cannot be, because it is always possible to change the class itself.
Every time the class changes, \texttt{find\_method} can potentially return a
new value.

@@

What is interesting here is that \texttt{\_find\_method} takes the \texttt{version}
argument but it does not use it at all. Its only purpose is to make the call
trace-elidable, because when the version object changes, the result of the call might be
different from the previous one.

@@

SPUR \cite{bebenita_spur:_2010} is a tracing JIT for CIL bytecode, which is then
used to trace through a JavaScript implementation written in C\#. The
JavaScript implementation compiles JavaScript to CIL bytecode together with an
implementation of the JavaScript object model. The object model uses maps
and inline caches to speed up operations on objects. The tracer traces through
the compiled JavaScript functions and the object model. SPUR contains two hints
that can be used to influence the tracer: one to prevent tracing of a C\#
function and one to force unrolling of a loop (PyPy has equivalent hints, but
they were not described in this paper).

@@

We already explored promotion in other contexts, such as earlier versions of
PyPy's JIT.
%as well as a Prolog partial evaluator \cite{bolz_towards_2009}
Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.
